Linear analysis of the vertical shear instability in astrophysical disks

### 2015-02-10 application.tex
first rough draft
spell-checked first rough draft

In order to utilize our linear results, which assumes $t_c$ is a
constant, we need to make further 
approximations. Specifically, we consider vertically-isothermal disks
and replace explicit $\rho$ depedences in the above expressions by $\Sigma/
H_\mathrm{iso}$, where $\Sigma$ is the surface density at the fiducial
radius of interest. For example, the dimensionless diffusion
coefficient becomes 
\begin{align}\label{diff_coeff_dimensionless}
  \hat{\eta} \to  \frac{16\sigma_s T^3}{3\kappa_d\Sigma^2
    c_v\Omega_k}. 
\end{align}
Furthermore, we will choose a temperature-dependent opacity model for
simplicity, so that $\kappa_d=\kappa_d(T)$. The dimensionless thermal
relaxation time $\beta$  
\begin{align}\label{real_beta}
  \beta \equiv t_c\Omega_k =
  \frac{1}{\hat{\eta}}\left(\frac{l_\mathrm{rad}^2}{3H_\mathrm{iso}^2}
    + \frac{1}{\khat^2}\right)  \simeq
  \frac{1}{\hat{\eta}}\left(\frac{1}{3\kappa_d^2\Sigma^2} 
    + \frac{1}{\khat^2}\right)
\end{align}
is then a constant for vertically-isothermal disks, 
and we can directly apply previous results.

\subsection{Evaluation for a fiducial disk model}
We evaluate Eq. \ref{real_beta} for the protoplanetary disk model
described in \cite{chiang10} and summarized in Appendix \ref{mmsn}
(where an opacity model is also chosen). Defining $\zeta\equiv
r/\mathrm{AU}$, we find Eq. \ref{real_beta} becomes
\begin{align}\label{beta_mmsn}
\beta =
\frac{4.4\times10^5}{\mu(\gamma-1)}\left(\frac{\hat{\kappa}_d\hat{\Sigma}^2}{\hat{T}}\right)\zeta^{-57/14}\left(\frac{8.3\times10^{-9}}{\hat{\kappa}_d^2\hat{T}^4\hat{\Sigma}^2}\zeta^{33/7}+\frac{1}{\khat^2}\right).    
\end{align}
We also evaluate the criterion Eq. \ref{iso_vsi_cond} for the fiducial
disk model,  
\begin{align}\label{bcrit_mmsn}
  \beta_\mathrm{crit} = \frac{1.44\times10^{-2}}{(\gamma
    -1)}\left(\frac{\hat{T}}{\mu}\right)^{1/2}\zeta^{2/7}. 
\end{align}
We recall $\beta < \beta_\mathrm{crit}$ is neccessary for the
fundamental VSI when $\khat\gg 1$. %For $\khat\to0$, this
%becomes a sufficient condition (see Fig. \ref{relax_bcrit} and
%\S\ref{iso_vsi_beta_crit}).

In Fig. \ref{mmsn_bcrit_bcool} we plot Eq. \ref{beta_mmsn} and
Eq. \ref{bcrit_mmsn} for the Minimum Mass Solar Nebula (MMSN) by setting
the scalings $\hat{T}=\hat{\Sigma}=\hat{\kappa}_d=1$, with $\mu =
2.33$ and $\gamma=1.4$. The figure suggests that the fundamental VSI
operates in the outer parts of the MMSN at radii on 
the order of tens of AU. However, the thermal relaxation time $\beta$
is not much shorter than $\beta_\mathrm{crit}$, so the VSI may be 
weak.

@@

The most unstable mode occurs at $\khat\simeq30$ with growth rate 
\begin{align*}
  \nu =0.058\epsilon\Omega_k = 3.9\times10^{-3}\Omega_k.  
\end{align*}
This corresponds to a characteristic growth time $\sim
40$ orbits. However, since growth rates are weakly depedent on $\khat$ for
$10\lesssim \khat \lesssim 60$, this may be taken as a representative growth
time of the fundamental VSI in the MMSN at $r=50\mathrm{AU}$. The
eigenfunction for $\khat=30$ is shown in Fig. \ref{mmsn_eigenW}.

@@

\subsubsection{Vertically-depedent thermal relaxation} 
Note that for a vertically isothermal disk and opacity $\kappa_d(T)$, 
we can restore the $z$ dependence in $\beta$ (due to the $\rho$
dependence in  $\eta$, see Eq. \ref{eta_def}) by making the transformation
$\beta(\khat)\to\beta(\khat/\sqrt{2\pi}\hat{\rho})\equiv 
\tilde{\beta}(\khat,z)$ in  Eq. \ref{real_beta}, where
$\hat{\rho}=\rho/\rho(z=0)$.  We have have repeated the example
calculation above with $\beta\to\tilde{\beta}$. 
%We find growth rates
%were reduced, but at most by $\sim 15\%$ at $\khat=1$. Eigenfunctions
%were similar.

>>>>

In order to utilize our linear results, which assumes $t_c$ is a
constant, we need to make further 
approximations. Specifically, we consider vertically-isothermal disks
and replace explicit $\rho$ dependencies in the above expressions by $\Sigma/
H_\mathrm{iso}$, where $\Sigma$ is the surface density at the fiducial
radius of interest. For example, the dimensionless diffusion
coefficient becomes 
\begin{align}\label{diff_coeff_dimensionless}
  \hat{\eta} \to  \frac{16\sigma_s T^3}{3\kappa_d\Sigma^2
    c_v\Omega_k}. 
\end{align}
Furthermore, we will choose a temperature-dependent opacity model for
simplicity, so that $\kappa_d=\kappa_d(T)$. The dimensionless thermal
relaxation time $\beta$  
\begin{align}\label{real_beta}
  \beta \equiv t_c\Omega_k =
  \frac{1}{\hat{\eta}}\left(\frac{l_\mathrm{rad}^2}{3H_\mathrm{iso}^2}
    + \frac{1}{\khat^2}\right)  \simeq
  \frac{1}{\hat{\eta}}\left(\frac{1}{3\kappa_d^2\Sigma^2} 
    + \frac{1}{\khat^2}\right)
\end{align}
is then a constant for vertically-isothermal disks, 
and we can directly apply previous results.

\subsection{Evaluation for a fiducial disk model}
We evaluate Eq. \ref{real_beta} for the protoplanetary disk model
described in \cite{chiang10} and summarized in Appendix \ref{mmsn}
(where an opacity model is also chosen). Defining $\zeta\equiv
r/\mathrm{AU}$, we find Eq. \ref{real_beta} becomes
\begin{align}\label{beta_mmsn}
\beta =
\frac{4.4\times10^5}{\mu(\gamma-1)}\left(\frac{\hat{\kappa}_d\hat{\Sigma}^2}{\hat{T}}\right)\zeta^{-57/14}\left(\frac{8.3\times10^{-9}}{\hat{\kappa}_d^2\hat{T}^4\hat{\Sigma}^2}\zeta^{33/7}+\frac{1}{\khat^2}\right).    
\end{align}
We also evaluate the criterion Eq. \ref{iso_vsi_cond} for the fiducial
disk model,  
\begin{align}\label{bcrit_mmsn}
  \beta_\mathrm{crit} = \frac{1.44\times10^{-2}}{(\gamma
    -1)}\left(\frac{\hat{T}}{\mu}\right)^{1/2}\zeta^{2/7}. 
\end{align}
We recall $\beta < \beta_\mathrm{crit}$ is necessary for the
fundamental VSI when $\khat\gg 1$. %For $\khat\to0$, this
%becomes a sufficient condition (see Fig. \ref{relax_bcrit} and
%\S\ref{iso_vsi_beta_crit}).

In Fig. \ref{mmsn_bcrit_bcool} we plot Eq. \ref{beta_mmsn} and
Eq. \ref{bcrit_mmsn} for the Minimum Mass Solar Nebula (MMSN) by setting
the scaling $\hat{T}=\hat{\Sigma}=\hat{\kappa}_d=1$, with $\mu =
2.33$ and $\gamma=1.4$. The figure suggests that the fundamental VSI
operates in the outer parts of the MMSN at radii on 
the order of tens of AU. However, the thermal relaxation time $\beta$
is not much shorter than $\beta_\mathrm{crit}$, so the VSI may be 
weak.

@@

The most unstable mode occurs at $\khat\simeq30$ with growth rate 
\begin{align*}
  \nu =0.058\epsilon\Omega_k = 3.9\times10^{-3}\Omega_k.  
\end{align*}
This corresponds to a characteristic growth time $\sim
40$ orbits. However, since growth rates are weakly dependent on $\khat$ for
$10\lesssim \khat \lesssim 60$, this may be taken as a representative growth
time of the fundamental VSI in the MMSN at $r=50\mathrm{AU}$. The
eigenfunction for $\khat=30$ is shown in Fig. \ref{mmsn_eigenW}.

@@

\subsubsection{Vertically-dependent thermal relaxation} 
Note that for a vertically isothermal disk and opacity $\kappa_d(T)$, 
we can restore the $z$ dependence in $\beta$ (due to the $\rho$
dependence in  $\eta$, see Eq. \ref{eta_def}) by making the transformation
$\beta(\khat)\to\beta(\khat/\sqrt{2\pi}\hat{\rho})\equiv 
\tilde{\beta}(\khat,z)$ in  Eq. \ref{real_beta}, where
$\hat{\rho}=\rho/\rho(z=0)$.  We have repeated the example
calculation above with $\beta\to\tilde{\beta}$, but found similar
results. 
%We find growth rates
%were reduced, but at most by $\sim 15\%$ at $\khat=1$. Eigenfunctions
%were similar.

### 2015-02-10 linear_adia.tex
spell-checked first rough draft

One effect of introducing finite thermal relaxation is buoyancy. 
The $\zhat^2$ depedence in $\delta v_{z1}$ then makes sense because it
is most significant for large $\zhat$, i.e. away from the midplane
where the effect of buoyancy first appears as $t_c$ is increased.

@@

We now have an equation in the same form as that for isothermal
perturbations (Eq. \ref{iso_ode3}):
\begin{align}
  0 = g^{\prime\prime} - \hat{z}\left(A - 2\alpha\right)g^\prime +
  \left(B + \alpha\right)g.
\end{align}
We seek polynomial solutions 
\begin{align}
  g(\zhat) = \sum_{m=0}^M b_m \zhat^m,
\end{align}
which requires
\begin{align}
  B + \alpha = M\left(A-2\alpha\right).\label{adia_disp0}
\end{align}
% We remark that for $q=0$ and $\beta\to\infty$, Eq. \ref{adia_disp0} is
% the same as Eq. 54 in \cite{lubow93} when the low-frequency, Keplerian
% approximation is made in the latter. 
% Eq. \ref{nearly_iso_explicit} is same form
% as the governing equation for the adiabatic case
% (Eq. \ref{adia_iso4}). However, the coefficients $\widetilde{B}$ and $\widetilde{C}$
% have a complicated dependence on $\hat{\sigma}$. 
% Applying the same solution method as in \S\ref{adia_explicit}, we 
% find the adiabatic dispersion relation, 
This is a $6^\mathrm{th}$ order 
polynomial in $\hat{\sigma}$:  
\begin{align}
  0 = \sum_{l=0}^{6}c_l\hat{\sigma}^l,\label{relax_disp}
\end{align}
where the coefficients $c_l$ are given in Appendix \ref{relax_coeff}.
This dispersion relation generally requires a numerical
solution. However, for the fundamental mode ($M=0$) some analytical
progress can be made, which we consider next.

>>>>

One effect of introducing finite thermal relaxation is buoyancy. 
The $\zhat^2$ dependence in $\delta v_{z1}$ then makes sense because it
is most significant for large $\zhat$, i.e. away from the midplane
where the effect of buoyancy first appears as $t_c$ is increased.

@@

We now have an equation in the same form as that for isothermal
perturbations (Eq. \ref{iso_ode3}):
\begin{align}
  0 = g^{\prime\prime} - \hat{z}\left(A - 2\alpha\right)g^\prime +
  \left(B + \alpha\right)g.
\end{align}
We seek polynomial solutions 
\begin{align}
  g(\zhat) = \sum_{m=0}^M b_m \zhat^m,
\end{align}
which requires
\begin{align}
  B + \alpha = M\left(A-2\alpha\right).\label{adia_disp0}
\end{align}
% We remark that for $q=0$ and $\beta\to\infty$, Eq. \ref{adia_disp0} is
% the same as Eq. 54 in \cite{lubow93} when the low-frequency, Keplerian
% approximation is made in the latter. 
% Eq. \ref{nearly_iso_explicit} is same form
% as the governing equation for the adiabatic case
% (Eq. \ref{adia_iso4}). However, the coefficients $\widetilde{B}$ and $\widetilde{C}$
% have a complicated dependence on $\hat{\sigma}$. 
% Applying the same solution method as in \S\ref{adia_explicit}, we 
% find the adiabatic dispersion relation, 
This is a $6^\mathrm{th}$ order 
polynomial in $\hat{\sigma}$:  
\begin{align}
  0 = \sum_{l=0}^{6}c_l\hat{\sigma}^l,\label{relax_disp}
\end{align}
where the coefficients $c_l$ are given in Appendix \ref{relax_coeff}.
This dispersion relation generally requires a numerical
solution. However, for the fundamental mode ($M=0$) some analytic
progress can be made, which we consider next.

### 2015-02-10 linear_iso.tex
first rough draft
spell-checked first rough draft

In deriving Eq. \ref{vertiso_gov}, the vertical dependence of
$D=\kappa^2(z)-\sigma^2$ was ignored because the replacement
$D\to\Omega_k^2$ was made before eliminating variables in favor of
$\delta v_z$. In Appendix \ref{adia_improve} we show that making this
replacement after eliminating variables, which requires the vertical
derivative of $D$, adds unecessary complexity for thin disks.

@@

Note that we can generalize Eq. \ref{vertiso_gov_nondim} to other
(non-Keplerian) rotation profiles by redfining $\khat$
as $k_xc_s/\kappa$, $\hat{\sigma}$ as $\sigma/\kappa$, and 
$\epsilon$ as $c_s/r\kappa$.

@@

\subsubsection{Neccessity of vertical shear for
  instability}\label{integral_relation} 
We can establish a necessary condition for instability by multipling
Eq. \ref{iso_ode1} by $\rho W^*$ and integrate vertically from
$\zhat=\zhat_1$ to $z=\zhat_2$. We neglect boundary 
terms when integrating by parts, by assuming $W$ or
$W^\prime$ vanishes at the boundaries, or that the boundaries are 
sufficiently far away so that the boundary terms are negligible because of the
decaying background density with increasing height. Then,
\begin{align}
  &\hat{\sigma}^2\left(1+\hat{k}^2\right)\int_{\zhat_1}^{\zhat_2}\rho|W|^2d\zhat \notag\\
  &=\int_{\zhat_1}^{\zhat_2}\rho|W^\prime|^2d\zhat 
  +\ii \epsilon q \hat{k}\int_{\zhat_1}^{\zhat_2}\rho f(\zhat) W^*W^\prime d\zhat.\label{integral_relation1}
\end{align}
It follows that for instability ($\imag\hat{\sigma}>0$), it is neccessary to
have $q\neq0$ or more generally $d\Omega^2/dz\neq 0$, i.e. vertical shear.

@@

For the vertically isothermal disk we have $|f(\hat{z})| =
|\hat{z}|\left(1+\epsilon^2\hat{z}^2\right)^{-3/2}$, which maximises
at $|\zhat|=1/\sqrt{2}\epsilon$. For vertical domains smaller 
than this height, the growth rate is limited by $|f|$ at the vertical
boundaries considered. For $|\epsilon\hat{z}|\ll1$ we have $f\simeq
\hat{z}$, in which case we can expect growth rates to increase at most linearly
with respect to the vertical domain size.   %no faster than linear

@@

\subsubsection{Polynomial solutions with vertical shear}\label{iso_poly}
We can seek power-series solution to Eq. \ref{iso_ode3},
\begin{align}
  W(\zhat) = \sum_{l=0}^\infty a_l\zhat^l. 
\end{align}
Then the coefficients must satisfy the recurrence relation
\begin{align}
  (l+2)(l+1)a_{l+2} +
  \left[n - l\left(1+\ii \epsilon q  \hat{k}\right)\right] a_l = 0, 
\end{align}
where $n$ is not neccessarily an integer. Indeed, if we demand
a polynomial of degree $L$ as the solution, then the eigenfrequency
must satisfy
\begin{align}\label{sig2_iso}
\hat{\sigma}^2 = L\left(\frac{1+\ii q \epsilon
    \hat{k}}{1+\hat{k}^2}\right).
\end{align}
The eigenfrequency $\hat{\sigma}$ is therefore complex for
$L\geq1$. We are interested in 
unstable modes for which $\imag{\hat{\sigma}}>0$ is the growth rate and is
given by 
\begin{align}\label{simple_growth}
  \hat{\nu} =\sqrt{
   \frac{L}{2\left(1+\hat{k}^2\right)}\left(\sqrt{1+q^2\epsilon^2\hat{k}^2} - 
    1\right)}. 
\end{align}
For fixed $q$ and $\epsilon$, the growth rate vanishes for both
$\hat{k}^2\to0$ and $\hat{k}^2\to\infty$. The maximum growth rate
occurs at the optimum wavenumber $\hat{k}_\mathrm{opt}$
\begin{align}
  |\hat{k}_\mathrm{opt}| = \sqrt{\frac{2+|\epsilon q|}{|\epsilon q|}},
\end{align}
and
\begin{align}
  \mathrm{max}\left(\hat{\nu}\right) =\frac{\sqrt{L}|\epsilon
    q|}{2\sqrt{1+|\epsilon q|}}. \label{iso_max_growth}
\end{align}
Since $L\geq1$, there is no limit to the maximum growth rate. This is
an artifact of the thin-disk approximation because the vertical 
shear $d\Omega^2/dz\propto z$ increases indefinitely with
height. However, we find Eq. \ref{iso_max_growth} compares well with
numerical solutions of the governing equation for small $L$.

>>>>

In deriving Eq. \ref{vertiso_gov}, the vertical dependence of
$D=\kappa^2(z)-\sigma^2$ was ignored because the replacement
$D\to\Omega_k^2$ was made before eliminating variables in favor of
$\delta v_z$. In Appendix \ref{adia_improve} we show that making this
replacement after eliminating variables, which requires the vertical
derivative of $D$, adds unnecessary complexity for thin disks.

@@

Note that we can generalize Eq. \ref{vertiso_gov_nondim} to other
(non-Keplerian) rotation profiles by redefining $\khat$
as $k_xc_s/\kappa$, $\hat{\sigma}$ as $\sigma/\kappa$, and 
$\epsilon$ as $c_s/r\kappa$.

@@

\subsubsection{Necessity of vertical shear for
  instability}\label{integral_relation} 
We can establish a necessary condition for instability by multiplying
Eq. \ref{iso_ode1} by $\rho W^*$ and integrate vertically from
$\zhat=\zhat_1$ to $z=\zhat_2$. We neglect boundary 
terms when integrating by parts, by assuming $W$ or
$W^\prime$ vanishes at the boundaries, or that the boundaries are 
sufficiently far away so that the boundary terms are negligible because of the
decaying background density with increasing height. Then,
\begin{align}
  &\hat{\sigma}^2\left(1+\hat{k}^2\right)\int_{\zhat_1}^{\zhat_2}\rho|W|^2d\zhat \notag\\
  &=\int_{\zhat_1}^{\zhat_2}\rho|W^\prime|^2d\zhat 
  +\ii \epsilon q \hat{k}\int_{\zhat_1}^{\zhat_2}\rho f(\zhat) W^*W^\prime d\zhat.\label{integral_relation1}
\end{align}
It follows that for instability ($\imag\hat{\sigma}>0$), it is necessary to
have $q\neq0$ or more generally $d\Omega^2/dz\neq 0$, i.e. vertical
shear. In Appendix \ref{global_vertiso} we reach a similar conclusion
in the global linear problem.

@@

For the vertically isothermal disk we have $|f(\hat{z})| =
|\hat{z}|\left(1+\epsilon^2\hat{z}^2\right)^{-3/2}$, which maximizes
at $|\zhat|=1/\sqrt{2}\epsilon$. For vertical domains smaller 
than this height, the growth rate is limited by $|f|$ at the vertical
boundaries considered. For $|\epsilon\hat{z}|\ll1$ we have $f\simeq
\hat{z}$, in which case we can expect growth rates to increase at most linearly
with respect to the vertical domain size.   %no faster than linear

@@

\subsubsection{Polynomial solutions with vertical shear}\label{iso_poly}
We can seek power-series solution to Eq. \ref{iso_ode3},
\begin{align}
  W(\zhat) = \sum_{l=0}^\infty a_l\zhat^l. 
\end{align}
Then the coefficients must satisfy the recurrence relation
\begin{align}
  (l+2)(l+1)a_{l+2} +
  \left[n - l\left(1+\ii \epsilon q  \hat{k}\right)\right] a_l = 0, 
\end{align}
where $n$ is not necessarily an integer. Indeed, if we demand
a polynomial of degree $L$ as the solution, then the eigenfrequency
must satisfy
\begin{align}\label{sig2_iso}
\hat{\sigma}^2 = L\left(\frac{1+\ii q \epsilon
    \hat{k}}{1+\hat{k}^2}\right).
\end{align}
The eigenfrequency $\hat{\sigma}$ is therefore complex for
$L\geq1$. We are interested in 
unstable modes for which $\imag{\hat{\sigma}}>0$ is the growth rate and is
given by 
\begin{align}\label{simple_growth}
  \hat{\nu} =\sqrt{
   \frac{L}{2\left(1+\hat{k}^2\right)}\left(\sqrt{1+q^2\epsilon^2\hat{k}^2} - 
    1\right)}. 
\end{align}
For fixed $q$ and $\epsilon$, the growth rate vanishes for both
$\hat{k}^2\to0$ and $\hat{k}^2\to\infty$. The maximum growth rate
occurs at the optimum wavenumber $\hat{k}_\mathrm{opt}$
\begin{align}
  |\hat{k}_\mathrm{opt}| = \sqrt{\frac{2+|\epsilon q|}{|\epsilon q|}},
\end{align}
and
\begin{align}
  \mathrm{max}\left(\hat{\nu}\right) =\frac{\sqrt{L}|\epsilon
    q|}{2\sqrt{1+|\epsilon q|}}. \label{iso_max_growth}
\end{align}
Since $L\geq1$, there is no limit to the maximum growth rate. This is
an artifact of the thin-disk approximation because the vertical 
shear $d\Omega^2/dz\propto z$ increases indefinitely with
height. However, we find Eq. \ref{iso_max_growth} compares well with
numerical solutions of the governing equation for small $L$.

### 2015-02-10 numerical.tex
spell-checked first rough draft

\subsection{VSI in a locally isothermal, neutrally stratified disk}\label{vertiso_pertiso} 
% disk model is a nearly-vertically
% isothermal disk with $\Gamma=1.011$, $p=-1.5$, $q=-1$,
% $\epsilon=0.1$. 
We first calculate the VSI in a nearly vertcally isothermal, neutrally
straitified disk ($N_z^2\equiv0$) by setting
$\gamma=\Gamma=1.011$ and $\beta=10^{-3}$. Other disk parameters are 
$(p,q,\epsilon)=(-1.5,-1,0.1)$ and the vertical domain is $\zmax =
5\epsilon r$. This setup is similar to that adopted in
\cite{mcnally14}, who calculated the VSI in vertically isothermal
disks subject to isothermal perturbations ($\Gamma=\gamma=1$).  This
allows us to test our numerical code.

@@

In Fig. \ref{lowfreq_eigen} we plot the eigenvalues 
$\sigma = \omega + \ii\nu$ found for $\khat=20\pi$. This plot is
similar to Fig. 3 in \cite{mcnally14}. The fundamental
VSI mode, with one node in $W$, has the smallest $|\sigma|$. The
eigenvalue calculated from Eq. \ref{sig2_iso} with $L=1$ and the 
numerically-obtained value are  
\begin{align*}
  (\omega,\nu) = (-0.3053,0.2606)\epsilon\Omega_k &\quad \text{(from
    Eq. \ref{sig2_iso})},\\
  (\omega,\nu) = (-0.3010,0.2575) \epsilon \Omega_k &\quad \text{(numerical)}.
\end{align*}
This agreement is surpsingly good given the number of approximations
used to obtain Eq. \ref{sig2_iso}, which also imposes a different
boundary condition to that in the numerical calculation. These values
are also close to that found by \cite{mcnally14}. The eigenfrequency
is not sensitive to the vertical boundary condition because the
expression for $\sigma$, given by Eq. \ref{integral_relation1},
involve integrands with $\rho$ as the   weight function, which rapidly
decays away from the midplane.

@@

For $\gamma>\Gamma$, i.e. stably stratified disks,
Fig. \ref{gcorr_compare_vnoniso} shows that introducing finite thermal
relaxation rapidly stabilizes the fundamental VSI mode. This is
similar to that observed for nearly vertically isothermal disks
(Fig. \ref{bcrit_compare1}). For $\gamma=1.7,\,2.0,\,2.5$, growth
rates reach zero for  $\beta\simeq0.17,\,0.083,\,0.045$, respectively.
Interestingly, these values are equal to
$\epsilon|q|/(\gamma-\Gamma)$. This suggests that the critial thermal
relaxation timescale for vertically non-isothermal disks can also be
estimated by Eq. \ref{iso_vsi_cond} but with $\gamma-1$ replaced by
$\gamma-\Gamma$.

>>>>

\subsection{VSI in a locally isothermal, neutrally stratified disk}\label{vertiso_pertiso} 
% disk model is a nearly-vertically
% isothermal disk with $\Gamma=1.011$, $p=-1.5$, $q=-1$,
% $\epsilon=0.1$. 
We first calculate the VSI in a nearly vertically isothermal, neutrally
stratified disk ($N_z^2\equiv0$) by setting
$\gamma=\Gamma=1.011$ and $\beta=10^{-3}$. Other disk parameters are 
$(p,q,\epsilon)=(-1.5,-1,0.1)$ and the vertical domain is $\zmax =
5\epsilon r$. This setup is similar to that adopted in
\cite{mcnally14}, who calculated the VSI in vertically isothermal
disks subject to isothermal perturbations ($\Gamma=\gamma=1$).  This
allows us to test our numerical code.

@@

In Fig. \ref{lowfreq_eigen} we plot the eigenvalues 
$\sigma = \omega + \ii\nu$ found for $\khat=20\pi$. This plot is
similar to Fig. 3 in \cite{mcnally14}. The fundamental
VSI mode, with one node in $W$, has the smallest $|\sigma|$. The
eigenvalue calculated from Eq. \ref{sig2_iso} with $L=1$ and the 
numerically-obtained value are  
\begin{align*}
  (\omega,\nu) = (-0.3053,0.2606)\epsilon\Omega_k &\quad \text{(from
    Eq. \ref{sig2_iso})},\\
  (\omega,\nu) = (-0.3010,0.2575) \epsilon \Omega_k &\quad \text{(numerical)}.
\end{align*}
This agreement is surprisingly good given the number of approximations
used to obtain Eq. \ref{sig2_iso}, which also imposes a different
boundary condition to that in the numerical calculation. These values
are also close to that found by \cite{mcnally14}. The eigenfrequency
is not sensitive to the vertical boundary condition because the
expression for $\sigma$, given by Eq. \ref{integral_relation1},
involve integrals with $\rho$ as the   weight function, which rapidly
decays away from the midplane.

@@

For $\gamma>\Gamma$, i.e. stably stratified disks,
Fig. \ref{gcorr_compare_vnoniso} shows that introducing finite thermal
relaxation rapidly stabilizes the fundamental VSI mode. This is
similar to that observed for nearly vertically isothermal disks
(Fig. \ref{bcrit_compare1}). For $\gamma=1.7,\,2.0,\,2.5$, growth
rates reach zero for  $\beta\simeq0.17,\,0.083,\,0.045$, respectively.
Interestingly, these values are equal to
$\epsilon|q|/(\gamma-\Gamma)$. This suggests that the critical thermal
relaxation timescale for vertically non-isothermal disks can also be
estimated by Eq. \ref{iso_vsi_cond} but with $\gamma-1$ replaced by
$\gamma-\Gamma$.

### 2015-02-10 paper.tex
spell-checked first rough draft

\begin{abstract}
  Baroclinic astrophysical disks with equilibrium rotation profiles
  $\Omega$ that depends on the distance away from the disk midplane are
  potentially unstable to the Vertical Shear Instability (VSI). It is
  is important to understand the linear instability before
  considering the VSI as a hydrodynamic route to turbulent angular
  momentum transport in accretion disks. We study the linear,
  axisymmetric VSI in radially-local but vertically-global models
  of accretion disks. We include an energy equation with thermal
  relaxation on a characteristic timescale $t_c=\beta\Omega_k^{-1}$,
  where $\beta$ is a dimensionless parameter and $\Omega_k$ is the
  Keplerian rotation frequency, and explore how the VSI depends on the
  thermodynamic response  of the disk by varying $\beta$. We find
  increasing $\beta$ away from zero dramatically reduces VSI growth
  rates because this generally introduces a strongly stabilizing
  vertical entropy gradient. Our linear calculations are consistent
  with previous hydrodynamic simulations, which indicates the VSI 
  requires a short thermal relaxation timscale. For vertically  
  isothermal disks, we quantify `fast'
  thermal relaxation as $\beta < \epsilon|q|/(\gamma-1)$, where
  $\epsilon$ is the disk aspect-ratio, $q$ is the radial power-law
  index for the equilibrium temperature profile (which gives rise to vertical 
  shear) and $\gamma$ is the adiabatic index. We apply our linear
  results to a disk model of the Minimum Mass Solar Nebulae (MMSN), and
  find the MMSN is unstable to the VSI at tens of AU with a
  characteristic growth times of tens of local orbital periods. We 
  suggest the VSI is a viable mechanism for angular momentum
  transport in the outer parts of protoplanetary disks. 
\end{abstract}

\section{Introduction}
% stuff about hydrodynamic angular momentum transport 
Understanding the mechanisms of angular mometum transport is the
central problem in the theory of accretion disks 
\citep{lyndenbell74}. The   
magneto-rotational instability in the context of differentially
rotating disks \citep[MRI,][]{balbus91} is now accepted as a robust
pathway to magneto-hydrodynamic (MHD) turbulence, which transports
angular momentum outwards and thus enables mass accretion.

@@

%recent work on vsi 
One possibiliy is purely hydrodynamic turbulence through the vertical
shear instability (VSI). The VSI is an axisymmetric linear instability
in three-dimensional (3D) disks in which the rotation frequency
$\Omega$ depends on the height from the disk midplane. The VSI is the
disk analog of the Goldreich-Schubert-Fricke instability
\citep[GSF,][]{goldreich67,fricke68} in differentially rotating stars.   
The height-dependence of $\Omega$, or vertical shear, enables
perturbations with large vertical lengthscales (relative to radial
lengthscales) to become unstable.    
\cite{umurhan13} gives a simple explanation for this 
instability: an exchange of fluid parcels while conserving their
angular momenta individually, may result in a new configuration of
lower energy, if the exchange occurs over a sufficiently large
vertical displacement.

@@

%what we're doing
In this work we study the linear VSI in radially-local,
vertically-global models of astrophysical disks. We include an energy
equation with a source term that allows the thermodynamic response of
perturbations between varied from isothermal to adiabatic. Our
formulation generalizes previous linear calculations of the 
VSI carried out by \cite{nelson13} and \cite{mcnally14} which assume
isothermal perturbations. Our models are also an extension of early
analysis of \cite{urpin98} and \cite{urpin03} who consider
vertically-localized disturbances. Our vertically-global approach is
necessary to study disturbances with vertical wavelengths
comparable to or larger than the disk scaleheight. We thus focus on
vertically-global perturbations, which were found by \cite{nelson13}
to dominate their numerical simulations.

>>>>

\begin{abstract}
  Baroclinic astrophysical disks with equilibrium rotation profiles
  $\Omega$ that depends on the distance away from the disk midplane are
  potentially unstable to the Vertical Shear Instability (VSI). It is
  is important to understand the linear instability before
  considering the VSI as a hydrodynamic route to turbulent angular
  momentum transport in accretion disks. We study the linear,
  axisymmetric VSI in radially-local but vertically-global models
  of accretion disks. We include an energy equation with thermal
  relaxation on a characteristic timescale $t_c=\beta\Omega_k^{-1}$,
  where $\beta$ is a dimensionless parameter and $\Omega_k$ is the
  Keplerian rotation frequency, and explore how the VSI depends on the
  thermodynamic response  of the disk by varying $\beta$. We find
  increasing $\beta$ away from zero dramatically reduces VSI growth
  rates because this generally introduces a strongly stabilizing
  vertical entropy gradient. Our linear calculations are consistent
  with previous hydrodynamic simulations, which indicates the VSI 
  requires a short thermal relaxation timescale. For vertically  
  isothermal disks, we quantify `fast'
  thermal relaxation as $\beta < \epsilon|q|/(\gamma-1)$, where
  $\epsilon$ is the disk aspect-ratio, $q$ is the radial power-law
  index for the equilibrium temperature profile (which gives rise to vertical 
  shear) and $\gamma$ is the adiabatic index. We apply our linear
  results to a disk model of the Minimum Mass Solar Nebulae (MMSN), and
  find the MMSN is unstable to the VSI at tens of AU with a
  characteristic growth times of tens of local orbital periods. We 
  suggest the VSI is a viable mechanism for angular momentum
  transport in the outer parts of protoplanetary disks. 
\end{abstract}

\section{Introduction}
% stuff about hydrodynamic angular momentum transport 
Understanding the mechanisms of angular momentum transport is the
central problem in the theory of accretion disks 
\citep{lyndenbell74}. The   
magneto-rotational instability in the context of differentially
rotating disks \citep[MRI,][]{balbus91} is now accepted as a robust
pathway to magneto-hydrodynamic (MHD) turbulence, which transports
angular momentum outwards and thus enables mass accretion.

@@

%recent work on vsi 
One possibility is purely hydrodynamic turbulence through the vertical
shear instability (VSI). The VSI is an axisymmetric linear instability
in three-dimensional (3D) disks in which the rotation frequency
$\Omega$ depends on the height from the disk midplane. The VSI is the
disk analog of the Goldreich-Schubert-Fricke instability
\citep[GSF,][]{goldreich67,fricke68} in differentially rotating stars.   
The height-dependence of $\Omega$, or vertical shear, enables
perturbations with large vertical lengthscales (relative to radial
lengthscales) to become unstable.    
\cite{umurhan13} gives a simple explanation for this 
instability: an exchange of fluid parcels while conserving their
angular momenta individually, may result in a new configuration of
lower energy, if the exchange occurs over a sufficiently large
vertical displacement.

@@

%what we're doing
In this work we study the linear VSI in radially-local,
vertically-global models of astrophysical disks. We include an energy
equation with a source term that allows the thermodynamic response of
perturbations between varied from isothermal to adiabatic. Our
formulation generalizes previous linear calculations of the 
VSI carried out by \cite{nelson13} and \cite{mcnally14} which assume
isothermal perturbations. Our models are also an extension of early
analysis of \cite{urpin98} and \cite{urpin03} who consider
vertically-localized disturbances. Our vertically-global approach is
necessary to study disturbances with vertical wavelengths
comparable to or larger than the disk scale height. We thus focus on
vertically-global perturbations, which were found by \cite{nelson13}
to dominate their numerical simulations.
